\section{Conclusion \& Outlook\vspace{-3mm}\label{sec:outlook}}
We have presented a framework for reasoning in rule-based WSML
that builds on a mapping to Datalog and on querying a generic
Datalog layer. The single well-defined transformation steps can be
reused across various adaptations for different scenarios in a
highly modular way. We have incorporated debugging features by
replacing native constraints with rules to derive
debugging-relevant information that can be queried by an ontology
engineer. We have implemented our framework with two existing
reasoner tools, namely KAON2 and MINS, as alternative
implementations of the generic Datalog layer, by which we provide
the first available reasoning system for the WSML language.

While the current framework focuses on WSML-Core, -Flight and
-Rule, efforts are ongoing to extend the transformations to
disjunctive Datalog and description logics. The KAON2 system
natively supports disjunctive Datalog and DL reasoning, the latter
even extended by WSML-Flight-like rules. Also the DLV
system~\cite{citrigno97dlv} (implementing disjunctive Datalog
under the stable model semantics) can be used to realise a similar
reasoning. Furthermore, we plan to integrate the KRHyper
system~\cite{wernhard03system}, which allows reasoning with
disjunctive logic programs with stratified default negation.
Transformations to DL additionally allow to incorporate
description logic system APIs to support efficient reasoning with
WSML-DL.


%\paragraph{Acknowledgements.}
%This work is supported by the European Commision under the the DIP
%project (FP6-507483), and by the Austrian Federal Ministry for
%Transport, Innovation, and Technology under the project {\sffamily
% {\bfseries R}W$^{\mathsf{2}}$} (FFG 809250).
