\def\transdebug{\transtxt{debug}}
\def\pvotype{\predicate{\predsubtxt{v\_otype}}}
\def\pvmincard{\predicate{\predsubtxt{v\_mincard}}}
\def\pvmaxcard{\predicate{\predsubtxt{v\_maxcard}}}
\def\pvuser{\predicate{\predsubtxt{v\_user}}}
\def\axiomid{\ensuremath{Ax_{I\!D}}\xspace}
\def\debugaxioms{\ensuremath{P_{\smtxtit{debug}}}\xspace}

\section{Debugging Support\label{sec:debugging}}
%-- briefly motivate debugging for the ontology engineering process \\
%-- -- during the process of ontology engineering, a modeller easily produces erroneous contradictory information, which he needs to get aware of \\
%-- -- the source for contradictory information in rule-based WSML are primarily constraints \\
%-- -- with plain datalog mechanisms, modellers only get aware of whether some constraint is violated, i.e. whether the empty clause is derived from $P_O$ indicating that the original ontology $O$ contains erroneous modelling; the modeller is not informed about the type of problem and why the problem occurs \\
%-- -- our framework offers debugging features that allow for identifying violated constraints and involved WSML entities, which is achieved by replacing constraints by rules with special additional meta-level predicates in their head \\

During the process of ontology development, an ontology engineer
can easily construct an erroneous model containing contradictory
information. In order to produce consistent ontologies,
inconsistencies should be reported to engineers with some details
about the ontological elements that cause the inconsistency.

In rule-based WSML, the source for erroneous modelling are always
constraints, together with a violating situation of concrete
instances related via attributes. The plain Datalog mechanisms
employed in the reasoning framework according to Section
\ref{sec:mapping} only allow for checking whether some constraint
is violated, i.e.\ whether the empty clause is derived from $P_O$
indicating that the original ontology $O$ contains errors -- more
detailed information about the problem is not reported. Experience
shows that it is a very hard task to identify and correct errors
in the ontology without such background information.

In our framework, we support debugging features that provide
information about the ontology entities which are involved in a
constraint violation. We achieve this by replacing constraints
with appropriate rules that contain the needed additional
information in their heads.

\subsection{Identifying Constraint Violations}
%-- describe the kind of debugging features the framework supports and what they allow for \\
%-- -- in a WMSL ontology, constraints can be violated by the instance situation, making the ontology inconsistent \\
%-- -- two things a of interest to the modeller when a constraint is violated: a) the type of constraint and the entities involved \\
%-- -- $<$ give an example of a violated constraint to illustrate what information is relevant for the modeller in such a situation $>$ \\
%-- -- in the different situations of violation of different types of constraints, different information is relevant for the modeller to repair the erroneous situation  \\
%-- -- $<$ list different types of violations $>$ \\

In case of an inconsistent ontology due to a constraint violation,
two things are of interest to the ontology engineer: a) the type
of constraint that is violated and b) the entities, i.e.\
concepts, attributes, instances, etc., that are involved in the
violation.

To give an example, consider the WSML ontology in Section
\ref{sec:wsml}. There, the attribute \syn{hasOnlineService} of the
concept \syn{ITBundle} is constrained to instances of type
\syn{OnlineService}. Suppose we replace the current value of the
attribute \syn{hasOnlineService} for the instance \syn{MyBundle}
by the instance \syn{MSNDialup}. Then, this constraint would be
violated because \syn{MSNDialup} is not an instance of the concept
\syn{OnlineService}. For an ontology engineer who needs to repair
this erroneous modelling, it is important to know the entities
that cause the violation, which in this case are the attribute
\syn{hasOnlineService} together with the range concept
\syn{OnlineService} and the non-conforming instance
\syn{MSNDialup}.

For the various types of constraint violations, the information
needed by the ontology engineer to track down the problem
successfully is different from case to case.

\paragraph{Attribute Type Violation} -- An attribute type
constraint of the form $C[a$ \wsml{ofType} $T]$ is violated
whenever an instance of the concept $C$ has value $V$ for the
attribute $a$, and it cannot be inferred that $V$ belongs to the
type $T$. Here, $T$ can be either a concept or a datatype, while
$V$ is then an instance or a data value, accordingly. In such a
situation, an ontology engineer is particularly interested in the
instance $I$, in the attribute value $V$ that caused the
constraint violation, together with the attribute $a$ and the
expected type $T$ which the value $V$ failed to adhere to.

\paragraph{Minimum Cardinality Violation} -- A minimum
cardinality constraint of the form \wsml{concept} $C$ $a \; (n$
\wsml{*}$)$, is violated whenever the number of distinguished
values of the attribute $a$ for some instance $I$ of the concept
$C$ is less than the specified cardinality $n$. In such a
situation, an ontology engineer is particularly interested in the
instance $I$ that failed to have a sufficient number of attribute
values, together with the actual attribute $a$. (Information about
how many values were missing can be learned by querying the
ontology separately.)

\paragraph{Maximum Cardinality Violation} -- A maximum
cardinality constraint of the form \wsml{concept} $C$ $a \;
($\wsml{0} $n)$, is violated whenever the number of distinguished
values of the attribute $a$ for some instance $I$ of the concept
$C$ exceeds the specified cardinality $n$. Again, here an ontology
engineer is particularly interested in the instance $I$ for which
the number of attribute values was exceeded, together with the
actual attribute $a$.

\paragraph{User-Defined Constraint Violation} -- Not only
built-in WSML constraints, but also user-defined constraints,
contained in an axiom definition of the form \wsml{axiom} \axiomid
\wsml{definedBy} $B$, can be violated. In this case, the
information which helps an ontology engineer to repair an
erroneous situation is dependent on the arbitrarily complex body
$B$ and cannot be determined in advance. However, a generic
framework can at least identify the violated constraint by
reporting the identifier \axiomid of the axiom.

To give an example, consider again the ontology from Section
\ref{sec:wsml}. Replacing the network connection \syn{ArcorDSL} of
\syn{MyBundle} by the slower \syn{MSNDialup} one results in the a
violation of the user-defined constraint specified by the axiom
named \syn{SharePriceFeed\_requires\_bandwidth}. This constraint
requires a certain bandwidth for connections in bundles with share
price feed online services, which is not met by \syn{MSNDialup},
and thus the ontology engineer is reported the axiom name that
identifies the violated constraint.


\subsection{Debugging by Meta-Level Reasoning}
%-- describe how these debugging features are realised via additional meta-level predicates an additional fixed set of rules \\
%-- -- in our framework we realise the debugging features for identifying constraint violations together with involved entities by replacing constraints with rules \\
%-- -- these rules have additional debugging-specific meta-level predicates in their heads which are instantiated when a constraint body evaluates to true; this way the debugging information is derived by datalog rules and can be queried for \\
%-- -- the replacements of constraints is included in the transformation pipeline $\tau$ as an additional step \\
%\begin{displaymath}
%    \tau = \transdlog \circ \translt \circ \transnorm \circ \transdebug \circ \transax
%\end{displaymath}
%-- -- the additional transformation step $\transdebug$ is applied after conceptual syntax has been resolved, replacing constraints on the level of WSML logical expressions \\
%-- -- the detailed constrained replacement performed by \transdebug can be seen from Table \ref{tab:debugging} \\
%-- -- the body variables are supposed to match the appropriate form of constraint body; notice: the semantics of ofType is encoded in the meta-level axioms \mlaxioms, so ofType-constraints can't be as easily replaced but have to be generated by \transdebug \\
%-- -- to maintain the constraint-semantics, some additional debugging-specific meta-level axioms, denoted by \debugaxioms, have to be included, which are shown in Table \ref{tab:debugging-axioms} \\
%-- -- thus, the datalog program used for reasoning with the original WSML ontology turns to: \\
%\begin{displaymath}
%    P_O = \mlaxioms \cup \debugaxioms \cup \tau(O)
%\end{displaymath}
%-- -- then one can ask for occurrences of the different kinds of constraint violation by e.g. \\
%\begin{displaymath}
%    \{(a,T,I,V) : (P_O , ?\pvotype(a,T,I,V)) \rightarrow \top \}
%\end{displaymath}
%-- -- which asks for all occurrences of type violations by means of datalog querying mechanisms; if this set is empty then there is no problem concerning types \\

In our framework, we realize the debugging features for reporting
constraint violations by replacing constraints with a special kind
of rules. Instead of deriving the empty clause, as constraints do,
these rules derive information about occurrences of constraint
violations by instantiating debugging-specific meta-level
predicates with the entities involved in a violation. In this way,
information about constraint violations can be queried for by
means of Datalog inferencing.

The replacement of constraints for debugging is included in the
transformation pipeline
\begin{displaymath}
    \tau = \transdpred \circ \transdlog \circ \translt \circ \transnorm \circ \transdebug \circ \transax
\end{displaymath}
where the additional transformation step \transdebug is applied
after the WSML conceptual syntax has been resolved, replacing
constraints on the level of WSML logical expressions. Table
\ref{tab:debugging} shows the detailed replacements performed by
\transdebug for the different kinds of constraints.

Minimal cardinality constraints (with bodies $B_{mincard}$) and
maximal cardinality constraints (with bodies $B_{maxcard}$) are
transformed to rules by keeping their respective bodies and adding
a head that instantiates one of the predicates \pvmincard and
\pvmaxcard to indicate the respective cardinality violation. The
variables for the involved attribute $a$ and instance $I$ are the
ones that occur in the respective constraint body $B$.

Similarly, a user-defined constraint is turned into a rule by
keeping the predefined body $B_{user}$ and including a head that
instantiates the predicate \pvuser to indicate a user-defined
violation. The only argument for the predicate \pvuser is the
identifier \axiomid of the axiom, by which the constraint has been
named.

Constraints on attribute types are handled differently because
these constraints are not expanded during the transformation
\transax ; they are rather represented by WSML
\wsml{ofType}-molecules for which the semantics is encoded in the
meta-level axioms \mlaxioms. In order to avoid the modification of
\mlaxioms in the reasoning framework, such molecules are expanded
by \transdebug, as shown in Table
\ref{tab:debugging}.\footnote{After this expansion of
\wsml{ofType} molecules, the respective axiom (4) in \mlaxioms for
realising the semantics of attribute type constraints does not
apply anymore.}

\medskip

To maintain the constraining semantics of the replaced
constraints, an additional set of meta-level axioms $\debugaxioms
\in \mathcal{P}$ is included for reasoning. The rules in
\debugaxioms derive the empty clause for any occurrence of a
constraint violation, as shown in Table
\ref{tab:debugging-axioms}.

Including the debugging features, the Datalog program for
reasoning about the original ontology then turns to
\begin{displaymath}
    P_O = \mlaxioms \cup \dataaxioms \cup \debugaxioms \cup \tau(O) \;\;\;.
\end{displaymath}
Occurrences of constraint violations can be recognized by querying
$P_O$ for instantiations of the various debugging-specific
meta-level predicates \pvotype, \pvmincard, \pvmaxcard and
\pvuser. For example, the set
\begin{displaymath}
    (P_O , \qury \pvotype(a,T,I,V))
\end{displaymath}
contains tuples for all occurrences of attribute type violations
in $P_O$, identifying the respective attribute $a$, expected type
$T$, involved instance $I$ and violating value $V$ for each
violation. This set is empty if there are no attribute types
violated.

\begin{table}[tb]\label{tab:debugging}\centering
\begin{footnotesize}
\begin{tabular}{|l|l|}
  \hline
  \rule{0cm}{3.2mm} {\normalsize \emph{Constraint}} & {\normalsize \emph{Rule}} \\
  \hline
  $\transdebug(\{\mve1, \dots , \mven\})$ & $\{\transdebug(\mve1), \dots , \transdebug(\mven)\}$ \\
  $\transdebug($\wsml{\cstr}$B_{mincard}.)$ & $\pvmincard(a,I)$\wsml{\lprl}$B_{mincard}.$ \\
  $\transdebug($\wsml{\cstr}$B_{maxcard}.)$ & $\pvmaxcard(a,I)$\wsml{\lprl}$B_{maxcard}.$ \\
  $\transdebug($\wsml{\cstr}$B_{user}.)$ & $\pvuser(\axiomid)$\wsml{\lprl}$B_{user}.$ \\
  $\transdebug(C[a$ \wsml{ofType} $T].)$ & $\pvotype(a,T,I,V)$\wsml{\lprl} \\
  & $\;C[a$ \wsml{ofType} $T]$ \wsml{and} $I$ \wsml{memberOf} $C$ \\
  & $\;I[a$ \wsml{hasValue} $V]$ \wsml{and naf} $V$\wsml{memberOf} $T.$ \\
  \hline
\end{tabular}
\end{footnotesize}
\caption{Replacing constraints by rules.}
\end{table}

\begin{table}[tb]\label{tab:debugging-axioms}\centering
\begin{small}
\begin{tabular}{|ll|}
  \hline
  \multicolumn{2}{|l|}{\rule{0cm}{3.2mm}{\normalsize \emph{Debugging Meta-Level Axioms}}} \\
  \hline
  (1) & $\dlogcstr \pvotype(a,T,I,V)$ \\
  (2) & $\dlogcstr \pvmincard(a,I)$ \\
  (3) & $\dlogcstr \pvmaxcard(a,I)$ \\
  (4) & $\dlogcstr \pvuser(\axiomid)$ \\
 \hline
\end{tabular}
\end{small} \caption{Meta-level axioms for debugging.}
\end{table}
